Inferring Community-scale Diversification from Genomic Data

### 2015-02-18 slides/bordeaux/slides-1.tex
Working on bordeaux slides.

\begin{document}
\maketitle

{
\setbeamercolor{background canvas}{bg=black}
\setbeamercolor{whitetext}{fg=white}
\begin{frame}
    \frametitle{Processes of diversification}
    \begin{columns}[c]
        \column{.5\textwidth}
        {\usebeamercolor[fg]{whitetext}
        \begin{itemize}
            \item<2-> Community-scale processes are important in
                biodiversification and community assembly
            \item<3-> Accounting for such processes will better our
                understanding of evolutionary history and
                community structure
            \item<4-> We need methods for inferring evolutionary patterns
                predicted by historical events from contemporary populations
        \end{itemize}
        }
        \column{.5\textwidth}
        \begin{figure}
            \begin{center}
            \includegraphics[width=\textwidth]{../images/earth-image.jpg}
            \end{center}
        \end{figure}
    \end{columns}
\end{frame}
}

@@

\begin{frame}[t]
    \frametitle<1-4>{Climate-driven diversification: Predictions}
    \frametitle<5->{Divergence-model choice}

    \centering{
    \includegraphics<1>[height=7.5cm]{../images/dmc-cartoon-islands.pdf}
    \includegraphics<2>[height=7.5cm]{../images/dmc-cartoon-taxa.pdf}
    \includegraphics<3>[height=7.5cm]{../images/dmc-cartoon-sea-level.pdf}
    \includegraphics<4>[height=7.5cm]{../images/dmc-cartoon-shared-no-labels.pdf}
    \includegraphics<5>[height=7.5cm]{../images/dmc-cartoon-shared.pdf}
    \includegraphics<6>[height=7.5cm]{../images/dmc-cartoon-2-1.pdf}
    \includegraphics<7>[height=7.5cm]{../images/dmc-cartoon-2-2.pdf}
    \includegraphics<8>[height=7.5cm]{../images/dmc-cartoon-2-3.pdf}
    \includegraphics<9>[height=7.5cm]{../images/dmc-cartoon-general.pdf}
    }
\end{frame}

\begin{frame}[t]
    \frametitle{Divergence-model choice}

@@

\begin{frame}
    \frametitle{New method: \dppmsbayes}
    \begin{itemize}
        % \item<1-> Reparameterized the model implemented in \msb
        \item<1-> Dirichlet-process prior (DPP) over all possible divergence
            models
        \item<2-> Flexible priors on continuous parameters to avoid strongly
            weighted posteriors
        \item<3-> Multi-processing interface to improve performance for
            accommodating genomic datasets
        % \item<4-> Uniform prior over divergence models
    \end{itemize}
    \barefootnote{\shortfullcite{Oaks2014dpp}}
\end{frame}

@@

\begin{frame}
    \frametitle{Current work: More data}
    \begin{itemize}
        \item<1-> Collecting genomic data from taxa co-distributed across Southeast
            Asia
        \item<2-> Preliminary results for 1000 loci from 5 pairs of \emph{Gekko
                mindorensis} populations
    \end{itemize}

@@

    \begin{itemize}[<+->]
        \item Full-likelihood Bayesian implementation
            \begin{itemize}
                \item Uses all the information in the data
                \item Applicable to deeper timescales
            \end{itemize}
        \item Analytically integrate over gene trees
            \footnote{\tiny\shortfullcite{Bryant2012}}
            \begin{itemize}
                \item Very efficient numerical approximation of posterior 
                \item Applicable to NGS datasets
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Climate-driven diversification model?}
    \begin{itemize}
        \item<1-> Results consistent with prediction of clustered
            divergences
        \item<2-> Results suggest multiple co-divergences
        \item<3-> However, there is a lot of uncertainty!
    \end{itemize}
    \barefootnote{\shortfullcite{Oaks2014dpp}}
\end{frame}

@@

\begin{frame}
    BIO 180 stuff
\end{frame}

\begin{frame}
    Software carpentry \\

    Workshops
\end{frame}

\begin{frame}
    \frametitle{Questions?}    
    \begin{center}
        {
        \Large
        \href{mailto:joaks1@gmail.com}{\tt joaks1@gmail.com}
        }
    \end{center}
\end{frame}

\begin{frame}
    \frametitle{Future directions---Computational}
    \begin{itemize}
        \item<1-> Full-phylogenetic framework
    \end{itemize}

    \begin{center}
        \includegraphics<1>[height=6.5cm]{../images/dmc-cartoon-no-islands-pop-shared-middle-wide.pdf}
        \includegraphics<2>[height=6.5cm]{../images/dmc-cartoon-no-islands-pop-phylogeny.pdf}
    \end{center}
\end{frame}

\begin{frame}
    \centerline{
    \includegraphics<1>[height=8cm]{../images/bd-cartoon-no-lines.pdf}
    \includegraphics<2>[height=8cm]{../images/bd-cartoon-shared.pdf}
    }
\end{frame}

\begin{frame}
\begin{columns}
    \column{.5\textwidth}
    \begin{minipage}[c]{\columnwidth}
    \begin{itemize}
        \item Blah
    \end{itemize}
    \end{minipage}

    \column{.6\textwidth}
    \begin{minipage}[t][\textheight][c]{\linewidth}
    \includegraphics<1>[width=0.75\columnwidth]{../images/bd-cartoon-shared-space.pdf}
    \includegraphics<2>[width=0.75\columnwidth]{../images/bd-cartoon-2-trees-shared.pdf}
    \end{minipage}
\end{columns}
\end{frame}

>>>>

\begin{document}

% \maketitle
\begin{frame}[noframenumbering]
    \titlepage
\end{frame}

{
\setbeamercolor{background canvas}{bg=black}
\setbeamercolor{whitetext}{fg=white}
\begin{frame}
    % \frametitle{Processes of diversification}
    \begin{columns}[c]
        \column{.5\textwidth}
        {\usebeamercolor[fg]{whitetext}
        \begin{itemize}
            \item<2-> Community-scale processes are important in
                biodiversification and community assembly
            \item<3-> Accounting for such processes will better our
                understanding of evolutionary history and
                community structure
            \item<4-> We need methods for inferring patterns predicted by
                community-scale events from contemporary populations
        \end{itemize}
        }
        \column{.5\textwidth}
        \begin{figure}
            \begin{center}
            \includegraphics[width=\textwidth]{../images/earth-image.jpg}
            \end{center}
        \end{figure}
    \end{columns}
\end{frame}
}

@@

\begin{frame}[t]
    % \frametitle<1-4>{Climate-driven diversification: Predictions}
    % \frametitle<5->{Divergence-model choice}
    \frametitle{Inferring community history}

    \centering{
    % \includegraphics<1>[height=7.5cm]{../images/dmc-cartoon-islands.pdf}
    % \includegraphics<2>[height=7.5cm]{../images/dmc-cartoon-taxa.pdf}
    \includegraphics<1>[height=7.5cm]{../images/dmc-cartoon-sea-level.pdf}
    % \includegraphics<4>[height=7.5cm]{../images/dmc-cartoon-shared-no-labels.pdf}
    \includegraphics<2>[height=7.5cm]{../images/dmc-cartoon-shared.pdf}
    \includegraphics<3>[height=7.5cm]{../images/dmc-cartoon-2-1.pdf}
    \includegraphics<4>[height=7.5cm]{../images/dmc-cartoon-2-2.pdf}
    \includegraphics<5>[height=7.5cm]{../images/dmc-cartoon-2-3.pdf}
    \includegraphics<6>[height=7.5cm]{../images/dmc-cartoon-general.pdf}
    }
\end{frame}

\begin{frame}[t]
    % \frametitle{Divergence-model choice}
    \frametitle{Inferring community history}

@@

\begin{frame}
    \frametitle{New method: \dppmsbayes}
    \begin{itemize}
        % \item<1-> Reparameterized the model implemented in \msb
        \item<1-> Dirichlet-process prior (DPP) over all possible divergence
            models
        \item<2-> Flexible priors on continuous parameters to avoid strongly
            weighted posteriors
        \item<3-> Multi-processing interface to accommodate genomic datasets
        % \item<4-> Uniform prior over divergence models
    \end{itemize}
    \barefootnote{\shortfullcite{Oaks2014dpp}}
\end{frame}

@@

\begin{frame}
    \frametitle{Results}
    \begin{itemize}
        \item<1-> Results consistent with prediction of clustered
            divergences
        \item<2-> Results suggest multiple interglacial periods were important
            drivers of diversification and community assembly across Islands of
            Southeast Asia
        \item<3-> However, there is a lot of uncertainty!
    \end{itemize}
    \barefootnote{\shortfullcite{Oaks2014dpp}}
\end{frame}

\begin{frame}[t]
    \frametitle{Current work: More data}
    \begin{minipage}[t][0.15\textheight][c]{\linewidth}
    \begin{itemize}
        \item<1-> Collecting genomic data from taxa co-distributed across
            Southeast Asian Islands and Mainland
        \item<2-> Preliminary results for 1000 loci from 5 pairs of \emph{Gekko
                mindorensis} populations
    \end{itemize}
    \end{minipage}

@@

    \begin{itemize}
        \item<1-> Full-likelihood Bayesian implementation
            \begin{itemize}
                \item<2-> Uses all the information in the data
                \item<2-> Applicable to deeper timescales
            \end{itemize}
        \item<3-> Analytically integrate over gene trees
            \footnote{\tiny\shortfullcite{Bryant2012}}
            \begin{itemize}
                \item<4-> Very efficient numerical approximation of posterior 
                \item<4-> Applicable to NGS datasets
            \end{itemize}
    \end{itemize}
\end{frame}

@@

{
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{../images/kane.jpg}}
\begin{frame}[t]
    \vspace{4.5cm}

    \uncover<2->{
    Working with the UW Biology Education Research Group (BERG) \\
    }

    \begin{itemize}
        \item<3-> Designing active-learning modules to teach fundamental concepts
            in evolution and statistics
        \item<4-> Taking an evidence-based approach to the classroom as a lecturer
            for Introductory Biology
    \end{itemize}
\end{frame}
}

\begin{frame}
    \frametitle{Collaborative teaching}
    \centerline{
    \includegraphics[height=1.5cm]{../images/software-carpentry.pdf}
    }
    Software Carpentry Bootcamp\\

    \begin{uncoverenv}<2->
    \begin{itemize}
        \item Two-day crash course in basic command line and software skills
        \item Taught an international group of biologists how to increase their
            research productivity.
    \end{itemize}
    \end{uncoverenv}

    Molecular Evolution Workshops\\

    \begin{uncoverenv}<3->
    \begin{itemize}
        \item Introduce international groups of graduate students, postdocs,
            and faculty to the fundamentals of molecular evolution and
            state-of-the-art software
    \end{itemize}
    \end{uncoverenv}
\end{frame}

\begin{frame}
    \frametitle{Acknowledgments}
    \begin{columns}[t]
        \column{.5\textwidth}
            {\bf Ideas and feedback:}
            \begin{myitemize}
                \item Leach\'{e} Lab
                \item Minin Lab
                \item Holder Lab
                \item Brown Lab/KU Herpetology
                \item Melissa Callahan
            \end{myitemize}
            \smallskip
            {\bf Computation:}\\
            \includegraphics<1->[height={8mm}]{../images/iplant.jpg}
            \hspace{0.5mm}
            \includegraphics<1->[height={8mm}]{../images/kuittc.png}
            \hspace{0.5mm}
            \includegraphics<1->[height={8mm}]{../images/uw.png}\\

        \column{.5\textwidth}
            {\bf Funding:}\\
            \includegraphics<1->[height={8mm}]{../images/nsf.jpg}
            \hspace{0.5mm}
            \includegraphics<1->[height={8mm}]{../images/ngs.jpg}
            \hspace{0.5mm}
            \includegraphics<1->[height={8mm}]{../images/ssb.png}\\

            \smallskip
            {\bf Photo credits:}
            \begin{myitemize}
                \item Rafe Brown, Cam Siler, Jesse Grismer, \& Jake Esselstyn
                \item FMNH Philippine Mammal Website:
                    \begin{myitemize}
                        \item D.S.\ Balete, M.R.M.\ Duya, \& J.\ Holden
                    \end{myitemize}
                \item PhyloPic!
            \end{myitemize}
    \end{columns}
\end{frame}

% \begin{frame}
%     \frametitle{Questions?}    
%     \begin{columns}[c]
%         \column{.5\textwidth}
%         \begin{center}
%             {
%             \Large
%             \href{mailto:joaks1@gmail.com}{\tt joaks1@gmail.com}
%             }
%         \end{center}
%         \column{.5\textwidth}
%             \begin{figure}
%                 \begin{center}
%                 \includegraphics[width=\textwidth]{../images/darwin-tol-copyright-boris-kulikov-2007.jpg}
%                 \caption{\tiny \copyright~2007 Boris Kulikov \href{http://boris-kulikov.blogspot.com/}{boris-kulikov.blogspot.com}}
%                 \end{center}
%             \end{figure}
%     \end{columns}
% \end{frame}

% Extra slides

\begin{frame}[noframenumbering]
    \frametitle{Easy as ABC}
    \centerline{
    \includegraphics<1>[width=\textwidth]{../images/rejection-sampling-observed.pdf}
    \includegraphics<2>[width=\textwidth]{../images/rejection-sampling-tolerance.pdf}
    \includegraphics<3>[width=\textwidth]{../images/rejection-sampling-10.pdf}
    \includegraphics<4>[width=\textwidth]{../images/rejection-sampling-100.pdf}
    \includegraphics<5>[width=\textwidth]{../images/rejection-sampling-200.pdf}
    \includegraphics<6>[width=\textwidth]{../images/rejection-sampling-500.pdf}
    \includegraphics<7>[width=\textwidth]{../images/rejection-sampling-1000.pdf}
    \includegraphics<8>[width=\textwidth]{../images/rejection-sampling-10000.pdf}
    \includegraphics<9>[width=\textwidth]{../images/rejection-sampling-20000.pdf}
    \includegraphics<10>[width=\textwidth]{../images/rejection-sampling-20000-post.pdf}
    }
\end{frame}

\begin{frame}[noframenumbering]
    \frametitle{Causes of bias: Insufficient sampling}
    \begin{itemize}
        \item Models with more parameter space are less densely sampled
        \item Could explain bias toward small models in extreme cases
        \item {\bf Predicts large variance in posterior estimates}
        \begin{itemize}
            \item We explored empirical and simulation-based analyses with
                2, 5, and 10 million prior samples, and estimates were
                very similar
        \end{itemize}
    \end{itemize}
    \smallskip
    \centerline{
    \includegraphics[width=\textwidth]{../images/omega_over_sampling.pdf}}
\end{frame}

\begin{frame}[noframenumbering]
    \frametitle{\dppmsbayes: Simulation results}
    % \vspace{1cm}
        \centerline{
        \includegraphics[width=\textwidth]{../images/old_old_power_psi_mode.pdf}}
        \vspace{0mm}
        \centerline{
        \includegraphics[width=\textwidth]{../images/old_u-shaped_power_psi_mode_headless.pdf}}
        \vspace{0mm}
        \centerline{
        \includegraphics[width=\textwidth]{../images/old_dpp_power_psi_mode_headless.pdf}}
    % \barefootnote{\shortfullcite{Oaks2014dpp}}
\end{frame}

\begin{frame}[noframenumbering]
    \frametitle{\dppmsbayes: Simulation results}
    % \vspace{1cm}
        \centerline{
        \includegraphics[width=\textwidth]{../images/old_old_power_omega_median.pdf}}
        \vspace{0mm}
        \centerline{
        \includegraphics[width=\textwidth]{../images/old_u-shaped_power_omega_median_headless.pdf}}
        \vspace{0mm}
        \centerline{
        \includegraphics[width=\textwidth]{../images/old_dpp_power_omega_median_headless.pdf}}
    % \barefootnote{\shortfullcite{Oaks2014dpp}}
\end{frame}

\begin{frame}[noframenumbering]
    \frametitle{Empirical results: Philippine diversification}
    \centerline{
    \includegraphics[width=0.8\textwidth]{../../empirical-analyses/plots/philippines-dpp-psi-posterior-old-vs-dpp-with-prior.pdf}}
    \barefootnote{\shortfullcite{Oaks2014dpp}}
\end{frame}
